Causal inference and the detection of coevolutionary contingencies using generalized dynamic phylogenetic models

### 2024-09-05 manuscript.Rmd
Update manuscript.Rmd

---
title: "Causal inference and the detection of coevolutionary contingencies using dynamic phylogenetic models"
author: "E. Ringen, J. S. Martin, S. Claessens, & A. V. Jaeggi"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: true
    includes:
      in_header: header.tex
bibliography: references.bib
link-citations: yes
linkcolor: blue
urlcolor: blue
figsintext : yes
---

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of
biological and cultural traits over evolutionary time. Among these methods, 
some of the most popular are phylogenetic generalised linear mixed models, 
phylogenetic path analysis, and Pagel's `discrete' method. While these methods
have clear benefits under specific scenarios, they also have important
limitations, such as unrealistic assumptions about the evolutionary process,
a focus on evolutionary correlation rather than causation, and limits on the 
number and type of traits that researchers can include. Here, we develop a
novel class of Bayesian \textit{dynamic phylogenetic models} (DPMs) that
overcomes these limitations and synthesizes the strengths of existing 
phylogenetic comparative methods into a single generalizable framework for 
dynamic phylogenetic causal inference. We formally and conceptually introduce
this class of model, which is applicable to any number of traits following any
response distribution. We provide a short tutorial for fitting these models
with the \texttt{coevolve} R package, a user-friendly package purpose-built for 
fitting DPMs in Stan. We highlight previous studies that have used DPMs to 
make inferences about trait coevolution, including analyses of brain size 
in non-human primates and social complexity in human societies, and suggest
some directions for future model development and research.
\end{center}

@@

We begin by briefly reviewing the benefits and constraints of current PCMs, 
particularly with regard to causal inference, exploring how their limitations 
motivate further methods development. We then formally and conceptually 
introduce a novel class of *dynamic phylogenetic models* (DPMs). We provide a 
worked synthetic example of coevolution between two continuous traits -- 
illustrated as female promiscuity and sperm length [@Fitzpatrick2009] -- to 
assess the accuracy and uncertainty of inferences made with dynamic phylogenetic
models across a reasonable range of sample sizes for phylogenetic research. We 
also provide an accompanying coding tutorial using our `coevolve` R package to
aid empiricists in easily applying basic dynamic phylogenetic models to their 
own datasets. We then demonstrate the generality and flexibility of our method 
with two empirical applications, which extend the model to complex, 
high-dimensional scenarios using dynamic latent variables. In particular, we 
show how our method can generate insights into the causal dynamics of trait 
coevolution across both genetic and cultural evolutionary timescales, using (1) 
a comparative dataset on primate brain size, sociality, diet, and life history 
traits [@DeCasien2017], which reduced to two coevolving latent variables, and 
(2) two studies on the evolution of social complexity across a global and a 
regional sample of pre-industrial human societies [@Ringen_preprint; 
@Sheehan2023].

@@

Fundamental to PCMs is the adjustment of raw trait associations for shared 
evolutionary history using a phylogenetic tree (or set of trees) and a 
statistical model. In a basic sense, phylogenetic adjustment is crucial for 
facilitating causal inference, as shared evolutionary history tends to generate 
trait correlations among closely related species with similar phenotypes, 
creating the illusion of convergent coevolution even when traits evolve 
independently. Adjustment for phylogeny reduces bias and variance due to 
clustering over the tree, thus reducing the risks of type I (false-positive), 
type II (false-negative), type M (magnitude), and type S (sign) errors during 
statistical inference (*CITE*). Nonetheless, adjustment for phylogeny is not a 
magic fix for all sources of unobserved confounding, nor does it guarantee that 
resulting estimates are causally interpretable [@Hansen2014]. Therefore, while all 
PCMs employ statistical techniques that reduce bias, these methods vary widely 
in the degree to which they isolate the causal pathways by which traits coevolve 
over time, with most commonly used methods focusing largely on evolutionary 
correlation (\autoref{fig:dags}a) rather than causation (\autoref{fig:dags}b-e).

@@

Phylogenetic generalized linear (mixed) models are linear regression models used
to quantify trait covariance between species or populations due to shared 
evolutionary history (likewise for their predecessor, independent contrasts) 
[@Blomberg2012; @Grafen1989; @Hadfield2010; @Lynch1991; @Symonds2014]. The most 
commonly assumed model of trait coevolution in these regression models is 
Brownian motion (also called a Wiener process, *CITE*), which assumes that the 
trait variance for a given species or population is proportional to the total 
branch length of the tree leading from root to tip, and the covariance between 
any two species is proportional to their amount of shared history. This process
is often biologically interpreted as reflecting patterns of evolutionary drift 
(or neutral evolution) across a phylogenetic tree. Empirically, the actual 
degree of covariance between related species or populations may be less than
expected under a pure Brownian motion model, such as when evolution is rapid or,
more mundanely, when measurement error is substantial. In these cases, the 
"phylogenetic signal" will be weak such that even closely related taxa may have 
dramatically different trait values [@Blomberg2003; @Kamilar2013], motivating 
various branch length transformations to the basic model to account for this 
greater independence among taxa [@Pagel1999]. In cases where traits do not 
evolve according to a modified Brownian motion process, other adjustments can 
also be made to the basic phylogenetic generalized linear model to modify the 
expected rate change in trait (co)variation. For instance, early burst models 
of adaptive radiation allow the rate of change to decrease over time (*CITE*), 
while Ornstein-Uhlenbeck models characterize evolutionary change as the product
of both stochastic and deterministic processes (*CITE*).

@@

To date, Pagel's [-@Pagel1994] "discrete" method is the only general application
of dynamic causal inference to phylogenetic data. By reconstructing the 
coevolutionary sequence of two traits along the branches of a phylogeny, 
researchers can infer causality and potentially reciprocal causality through 
temporal contingency, referred to as "Granger causality" in economics 
[@Granger1969]. For example, researchers can infer that $X$ evolved first and 
then made the evolution of $Y$ more likely. This method has resulted in many 
high-profile publications [@Cornwallis2017; @Fitzpatrick2009; @Kappeler2019; 
@Sheehan2018; @Shultz2011; @Watts2016], all of which are built on the stronger 
causal inference of temporal contingency.

@@

The limitation of Pagel's method to binary traits is further exacerbated when 
studying the (co)evolution of multiple continuous traits, which often motivates
the incorporation of latent variables into a dynamic causal analysis 
(\autoref{fig:dags}e). Latent variables are widely used in biology to capture
theoretically pertinent constructs that are difficult to directly operationalize
and quantify using a single measurement, such as size and shape dimensions in 
morphometrics (*CITE*), environmental quality and climate metrics in ecology
(*CITE*), life history variation [@Stott2024], and canonical axes of 
correlational selection in quantitative genetics (*CITE*). While latent 
variables are sometimes used pragmatically as placeholders for deeper causal 
analysis [@Shipley2016], many evolutionary models directly conceptualize and 
model causal effects occurring among latent variables. For instance, 
life-history traits are theorized to evolve along a latent fast-slow continuum 
[@Bielby2007; @Healy2019; @Stott2024; @Wright2019] that can both cause and in 
turn be caused by the correlated evolution of a niche complexity continuum 
reflecting differences in brain size and cognitive ability, social group size, 
and foraging behavior [@Dunbar2017; @Miller2019; @Navarrete2016] (see 
\autoref{sec:brain-size-evo}). While prior research has used methods such as 
phylogenetic principal component analysis (pPCA; *CITE*) to apply Pagel's method
to latent variables [e.g. @Cornwallis2017], this approach still requires falsely
dichotomizing latent variables and thus suffers from the same statistical biases
as with any other continuous measure. Synthesizing continuous and multivariate
methods with dynamic causal inference thus remains a major challenge for 
contemporary PCMs.

@@

The limitations of current PCMs can be overcome by synthesizing their respective
strengths into a single generalizable framework for dynamic phylogenetic causal
inference, applicable to both discrete and continuous traits. We have attempted
to do so through the development of a novel class of PCMs we refer to simply as
dynamic phylogenetic models (DPMs) of trait coevolution. Our DPM is driven by a 
multivariate stochastic differential equation, similar to a multivariate 
Ornstein-Uhlenbeck (OU) process (see Box \autoref{box:ou-model} for an overview
of the basic OU model) with greater model flexibility and complexity than is 
permitted by standard implementations. Our approach arose from two complementary
goals: first, to extend previous Bayesian implementations of the multivariate OU
model such that the optimal trait value $\theta$ is updated dynamically (rather
than assuming piecewise constant, see Box \autoref{box:ou-model}). Second, we 
wanted a general form for the optimal trait value $\theta$ that could be used to
assess directionality (X → Y vs. Y → X) and contingencies (X, then Y) in 
evolution, akin to Pagel's [-@Pagel1994] method but without restrictions on the
type or number of traits.

\mybox[label=box:ou-model]{The Ornstein-Uhlenbeck (OU) model}
The OU model is a mean-reverting, stationary Gauss-Markov process. It describes
change in a trait due to both Gaussian noise and reversion towards some central
value. In some fields, such as economics, the mean reversion component is called
"drift" and the stochastic noise is called "diffusion" (i.e., the Vasicek model,
*CITE*). In evolutionary biology, the mean-reverting quality is loosely 
interpreted as "selection" and accordingly the Gaussian noise is labeled "drift"
(*CITE*). The basic OU form is

where $\alpha$ controls the strength of mean-reverting selection, $\theta$ is 
the mean trait value, and $\sigma$ controls the strength of drift. When 
$\alpha = 0$, the model is pure drift Brownian motion. The simplest OU models 
assume a single evolutionary optimum ($\theta$), or estimate an ancestral 
optimum along with a global optimum. More elaborate OU models imagine that 
$\theta$ changes as a function of other variables, turning it into a 
coevolutionary process with varying selection regimes (i.e., the Hansen model; 
*CITE*). These approaches exploit the fact that, if selection regimes are 
piecewise-constant (the optimum is the same within each segment of the 
phylogenetic tree but allowed to vary at branching points), the OU process can 
be discretized, giving a closed form solution for the expectation and covariance
matrix of a trait. Butler and King (*CITE*) provided a maximum-likelihood 
algorithm for this approach, and a Bayesian implementation was developed by 
Ross et al. (*CITE*). Depending on the research question, the piecewise constant 
assumption can be quite restrictive and a poor approximation when tree segments
are long, due to the assumption of a constant optimum within each discrete tree
segment.

\endmybox

@@

where $\textbf{A}_{\text{\#}}=(\textbf{A}\otimes\textbf{I})+(\textbf{I}\otimes\textbf{A})$, 
with $\otimes$ denoting the Kronecker product. $\textbf{I}$ is an identity 
matrix, $\text{row()}$ is an operation that takes elements of a matrix row-wise
and puts them in a column vector, and $\text{irow()}$ is the inverse of the row
operation [@Driver2017]. In the OU process, $\frac{dy}{dt}=0$ when $y=\theta$ 
(Box \autoref{box:ou-model}). Equivalently, we can calculate the equilibrium 
value $\theta$ for each trait $\eta_i$ in the DPM as

@@

Note that the vector $\theta_{\eta}$ does not equal the time asymptotic trait 
values for the system as a whole ($\textbf{b}_{\Delta\infty}$; *CITE*), which 
are determined by

@@

\begin{algorithm}
  \caption{Implementation of the dynamic phylogenetic model}\label{alg:alg}
  \DontPrintSemicolon
  \For{$n \in N$}{
    Divide the evolutionary history of $n$ into $S$ segments, where each 
    segment starts with a parent node and ends with a child node or tip.\;
    Calculate the length of each segment $s \in S$, i.e., duration of time 
    between the parent and child nodes.\;
    Initialize the ancestral trait values $\eta_0$.\;
    \For{$s \in S$}{
      Solve for $\eta(s)$:
      \begin{equation}
        \nonumber
        \eta(s) = e^{\textbf{A}(t(s))}\eta(s-1)\textbf{A}^{-1}[e^{\textbf{A}
        (t(s)-t(s-1))}]\textbf{b} + \int_{t(s-1)}^{t(s)}e^{\textbf{A}
        (t(s-1)-t(s))}\textbf{G}dW(t(s))
      \end{equation}
      \;
    }
  }
\end{algorithm}

```{r dpm-algorithm, echo=FALSE, fig.align='center', fig.cap='\\textit{The DPM algorithm.} Example of how our phylogenetic algorithm works (see \\autoref{alg:alg} for formal structure). The evolutionary history of each tip (A-E) is described as a time series starting at \\(t(s)\\) and comprised of a set of segments \\(S\\). The OU process runs over each segment in order, where the initial trait values for any segment \\(s\\) are the terminal values from \\(s-1\\). Each sequence is a combination of shared segments (where tips are evolving together) and unshared segments (where tips are evolving independently). For example, S(A) and S(B) start with the same trait values because they both descend directly from the parent segment S(A,B). But further evolution along S(A) and S(B) happens independently. Dashed lines indicate segment split points.', out.width='100%'}
knitr::include_graphics("figures/dpm_algorithm.png")
```

@@

To aid researchers in applying the DPM to their own data, we have developed the 
`coevolve` R package (https://github.com/ScottClaessens/coevolve). This 
fully documented package provides a user-friendly interface for applying DPMs, 
with simple functions for generating DPM Stan code and data lists, fitting DPMs
to data using the `cmdstanr` package [@R_cmdstanr], and post-processing and 
plotting model results.

@@

We used the `SBC` R package [@Modrak2023] to run our simulation-based 
calibration exercise. We generated 500 simulated datasets from the prior ($N$ =
20, one continuous trait and one binary trait) and fitted the DPM to these
simulated datasets. We then determined whether the posterior ranks from the 
fitted models were uniformly distributed by plotting, for several key model 
parameters and the log-likelihood, the differences between the empirical 
cumulative distribution function and a perfectly uniform cumulative distribution
function. If posterior ranks are perfectly uniformly distributed as expected, 
this should result in a flat line, though some variation from flat is to be 
anticipated. Ideally, we would expect to see the lines falling within the 
expected deviations at the 95% level (blue ovals). Indeed, this is what we find
for our model (\autoref{fig:sbc}), suggesting that the Stan program is 
well-calibrated.

@@

Brain size, both in absolute terms and relative to body size, varies 
tremendously among taxa, with primates as a group having larger brains than 
other mammals, and humans having the largest brains of all primates [@Boddy2012;
@Isler2008; @Miller2019; @Smaers2021]. This has motivated a large number of 
comparative phylogenetic analyses on correlates of brain size, typically framed
as testing social or ecological benefits or constraints for evolving larger 
brains [@DeCasien2017; @Dunbar2017; @Isler2009]. However, many have recognized 
that this approach has reached an impasse, as different hypotheses can be 
supported depending on the specific dataset [@Powell2017; @Wartel2019] due to 
variable patterns of missingness and high collinearity among predictors (e.g. 
diet, social group size, and life history), and reciprocal causation is inherent
in most theoretical models (e.g. larger brains allow access to better diet which
removes energetic constraints on larger brains) [@Isler2014]. Add the fact that
Pagel's discrete method has to our knowledge never been applied to brain size 
due to its continuous nature, and this field is virtually predisposed to the use
of DPMs. The issues and innovations highlighted here should also generalize to 
many other morphological and socio-ecological traits.

To illustrate the potential of DPMs to analyze brain size evolution, we used the
dataset of @DeCasien2017, augmented by life-history variables from 
@Herculano2019, and a consensus phylogeny from 10ktrees [@Arnold2010]. We 
contrasted a typical phylogenetic multiple regression (Fig. 5a), in which all 
predictors are interpreted as independent causes of brain size, with a latent 
factor model (Fig. 5b), which was derived in a bottom-up way using exploratory 
graph analysis (see Supplementary Materials). Note that instead of modeling 
absolute brain size and adjusting for body size as is typical for phylogenetic
regression (Fig. 5a), the model in Fig. 5b directly modeled relative brain size
as the allometric slope for the brain-body relationship; we also accounted for
variable allometric relationships using random intercepts and slopes at the
species level, thus allowing species to differ in their brain-body allometry
[@Smaers2021]. Note also that this DPM used Bayesian imputation rather than 
removing species with missing data and losing valuable information and 
statistical power. In contrast to the approach in Fig. 5a, which compares 
independent causes and might conclude that "primate brain size is predicted by
diet but not sociality" [@DeCasien2017], the causal structure in Fig 5b groups
diet, social group size, and relative brain size into a single latent factor, 
termed here "niche complexity", with another latent factor, "life-history pace",
with loadings from the variables body size, age at maturity, and longevity. 
These latent factor groupings are not surprising, given that much work in 
primate socio-ecology has emphasized connections between diet and group size
[@Sterck1997; @Wrangham1980], and both are consistently found to correlate with
relative brain size [@DeCasien2017; @Dunbar2017]; the integration of life 
history traits and body size is also well established (*CITE*).

A DPM analyzing the coevolution of these two latent factors (see Supplementary
Materials) showed that evolutionary changes in life-history pace (LH) cause 
evolutionary changes in niche complexity (NC), but not vice versa (Fig. 5c). In
other words, evolution of larger size and slower life history was inferred to 
precede increases in relative brain size, group size, and frugivory. This is 
consistent with the observations that repeated changes in size characterized
mammalian evolution as lineages radiated into new niches [@Pagel2022] and 
changes in body size associated with new niches were arguably primarily 
responsible for changes in relative brain size [@Smaers2021]. Further, debates
about the relative importance of diet and sociality in brain size evolution 
[@DeCasien2017; @Dunbar2017] seem to be moot as both are sides of the same 
socio-ecological coin intimately tied to relative brain size [see also 
@Powell2017; @Wartel2019]. Many extensions of this model are possible, and more
theoretically motivated causal models should be explored to further advance our
understanding of brain size evolution.

>>>>

---
title: "Causal inference and the detection of coevolutionary contingencies using generalized dynamic phylogenetic models"
author: "E. Ringen, J. S. Martin, S. Claessens, & A. V. Jaeggi"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    number_sections: true
    includes:
      in_header: header.tex
bibliography: references.bib
link-citations: yes
linkcolor: blue
urlcolor: blue
figsintext : yes
---

\begin{center}
Phylogenetic comparative methods are widely used to study the coevolution of
biological and cultural traits over evolutionary time. Among these methods, 
some of the most popular are phylogenetic generalised linear mixed models, 
phylogenetic path analysis, and Pagel's `discrete' method. While these methods
have clear benefits under specific scenarios, they also have important
limitations, such as unrealistic assumptions about the evolutionary process,
a focus on evolutionary correlation rather than causation, and limits on the 
number and type of traits that researchers can include. Here, we develop a
novel class of Bayesian \textit{generalized dynamic phylogenetic models} (GDPMs) that
overcomes these limitations and synthesizes the strengths of existing 
phylogenetic comparative methods into a single generalizable framework for 
dynamic phylogenetic causal inference. We formally and conceptually introduce
this class of model, which is applicable to any number of traits following any
response distribution. We provide a short tutorial for fitting these models
with the \texttt{coevolve} R package, a user-friendly package purpose-built for 
fitting GDPMs in Stan. We highlight previous studies that have used GDPMs to 
make inferences about trait coevolution, including analyses of brain size 
in non-human primates and social complexity in human societies, and suggest
some directions for future model development and research.
\end{center}

@@

We begin by briefly reviewing the benefits and constraints of current PCMs, 
particularly with regard to causal inference, exploring how their limitations 
motivate further methods development. We then formally and conceptually 
introduce a novel class of *generalized dynamic phylogenetic models* (GDPMs). We provide a 
worked synthetic example of coevolution between two continuous traits -- 
illustrated as female promiscuity and sperm length [@Fitzpatrick2009] -- to 
assess the accuracy and uncertainty of inferences made with dynamic phylogenetic
models across a reasonable range of sample sizes for phylogenetic research. We 
also provide an accompanying coding tutorial using our `coevolve` R package to
aid empiricists in easily applying basic dynamic phylogenetic models to their 
own datasets. We then demonstrate the generality and flexibility of our method 
with two empirical applications, which extend the model to complex, 
high-dimensional scenarios using dynamic latent variables. In particular, we 
show how our method can generate insights into the causal dynamics of trait 
coevolution across both genetic and cultural evolutionary timescales, using (1) 
a comparative dataset on primate brain size, sociality, diet, and life history 
traits [@DeCasien2017], which reduced to two coevolving latent variables, and 
(2) two studies on the evolution of social complexity across a global and a 
regional sample of pre-industrial human societies [@Ringen_preprint; 
@Sheehan2023].

@@

Fundamental to PCMs is the adjustment of raw trait associations for shared 
evolutionary history using a phylogenetic tree (or set of trees) and a 
statistical model. In a basic sense, phylogenetic adjustment is crucial for 
facilitating causal inference, as shared evolutionary history tends to generate 
trait correlations among closely related species with similar phenotypes, 
creating the illusion of convergent coevolution even when traits evolve 
independently. Adjustment for phylogeny reduces bias and variance due to 
clustering over the tree, thus reducing the risks of type I (false-positive), 
type II (false-negative), type M (magnitude), and type S (sign) errors during 
statistical inference [@felsenstein1985phylogenies; @harvey1998phylogenetic]. Nonetheless, adjustment for phylogeny is not a 
magic fix for all sources of unobserved confounding, nor does it guarantee that 
resulting estimates are causally interpretable [@Hansen2014]. Therefore, while all 
PCMs employ statistical techniques that reduce bias, these methods vary widely 
in the degree to which they isolate the causal pathways by which traits coevolve 
over time, with most commonly used methods focusing largely on evolutionary 
correlation (\autoref{fig:dags}a) rather than causation (\autoref{fig:dags}b-e).

@@

Phylogenetic generalized linear (mixed) models are linear regression models used
to quantify trait covariance between species or populations due to shared 
evolutionary history (likewise for their predecessor, independent contrasts) 
[@Blomberg2012; @Grafen1989; @Hadfield2010; @Lynch1991; @Symonds2014]. The most 
commonly assumed model of trait coevolution in these regression models is 
Brownian motion, which assumes that the trait variance for a given species or population is proportional to the total 
branch length of the tree leading from root to tip, and the covariance between 
any two species is proportional to their amount of shared history [@felsenstein1985phylogenies]. This process
is often biologically interpreted as reflecting patterns of evolutionary drift 
(or neutral evolution) across a phylogenetic tree. Empirically, the actual 
degree of covariance between related species or populations may be less than
expected under a pure Brownian motion model, such as when evolution is rapid or,
more mundanely, when measurement error is substantial. In these cases, the 
"phylogenetic signal" will be weak such that even closely related taxa may have 
dramatically different trait values [@Blomberg2003; @Kamilar2013], motivating 
various branch length transformations to the basic model to account for this 
greater independence among taxa [@Pagel1999]. In cases where traits do not 
evolve according to a modified Brownian motion process, other adjustments can 
also be made to the basic phylogenetic generalized linear model to modify the 
expected rate change in trait (co)variation. For instance, early burst models 
of adaptive radiation allow the rate of change to decrease over time [@harmon2010early], 
while Ornstein-Uhlenbeck models characterize evolutionary change as the product
of both stochastic and deterministic processes [@Hansen2014; @butler2004phylogenetic].

@@

To date, Pagel's [-@Pagel1994] "discrete" method is the most widely-used application
of dynamic causal inference to phylogenetic data. By reconstructing the 
coevolutionary sequence of two traits along the branches of a phylogeny, 
researchers can infer causality and potentially reciprocal causality through 
temporal contingency, referred to as "Granger causality" in economics 
[@Granger1969]. For example, researchers can infer that $X$ evolved first and 
then made the evolution of $Y$ more likely. This method has resulted in many 
high-profile publications [@Cornwallis2017; @Fitzpatrick2009; @Kappeler2019; 
@Sheehan2018; @Shultz2011; @Watts2016], all of which are built on the stronger 
causal inference of temporal contingency.

@@

The limitation of Pagel's method to binary traits is further exacerbated when 
studying the (co)evolution of multiple continuous traits, which often motivates
the incorporation of latent variables into a dynamic causal analysis 
(\autoref{fig:dags}e). Latent variables are widely used in biology to capture
theoretically pertinent constructs that are difficult to directly operationalize
and quantify using a single measurement, such as size and shape dimensions in 
morphometrics [@zelditch2012geometric], environmental quality and climate metrics in ecology
[@arhonditsis2006exploring], life history variation [@Stott2024], and canonical axes of 
correlational selection in quantitative genetics [@posthuma2003theory]. While latent 
variables are sometimes used pragmatically as placeholders for deeper causal 
analysis [@Shipley2016], many evolutionary models directly conceptualize and 
model causal effects occurring among latent variables. For instance, 
life-history traits are theorized to evolve along a latent fast-slow continuum 
[@Bielby2007; @Healy2019; @Stott2024; @Wright2019] that can both cause and in 
turn be caused by the correlated evolution of a niche complexity continuum 
reflecting differences in brain size and cognitive ability, social group size, 
and foraging behavior [@Dunbar2017; @Miller2019; @Navarrete2016] (see 
\autoref{sec:brain-size-evo}). While prior research has used methods such as 
phylogenetic principal component analysis to apply Pagel's method
to latent variables [e.g. @Cornwallis2017], this approach still requires falsely
dichotomizing latent variables and thus suffers from the same statistical biases
as with any other continuous measure. Synthesizing continuous and multivariate
methods with dynamic causal inference thus remains a major challenge for 
contemporary PCMs.

## Dynamic methods with continuous traits

There are several other approaches to dynamic causal inference, based on the Ohrnstein-Uhlenbeck (OU) model, that relax Pagel's assumption of discrete traits. 
The OU model is a mean-reverting, stationary Gauss-Markov process. It describes
change in a trait due to both Gaussian noise and reversion towards some central
value. In some fields, such as economics, the mean reversion component is called
"drift" and the stochastic noise is called "diffusion" (i.e., the Vasicek model,
[@vasicek1977equilibrium]). In evolutionary biology, the mean-reverting quality is loosely 
interpreted as "selection" and accordingly the Gaussian noise is labeled "drift"
[@uhlenbeck1930theory; @hansen1997stabilizing; @Hansen2014]. The basic OU form is

where $\alpha$ controls the strength of mean-reverting selection, $\theta$ is 
the mean trait value, and $\sigma$ controls the strength of drift. When 
$\alpha = 0$, the model is pure drift Brownian motion. The simplest OU models 
assume a single evolutionary optimum ($\theta$), or estimate an ancestral 
optimum along with a global optimum. More elaborate OU models imagine that 
$\theta$ changes as a function of other variables, turning it into a 
coevolutionary process with varying selection regimes (i.e., the Hansen model; 
[@hansen1997stabilizing]). These approaches exploit the fact that, if selection regimes are 
piecewise-constant (the optimum is the same within each segment of the 
phylogenetic tree but allowed to vary at branching points), the OU process can 
be discretized, giving a closed form solution for the expectation and covariance
matrix of a trait. Butler and King [@butler2004phylogenetic] provided a maximum-likelihood 
algorithm for this approach, and a Bayesian implementation was developed by 
Ross et al. [@ross2016origins]. These approaches assume that one trait influences the evolution of another, 
but not vice versa (i.e. no feedback). The multivariate OU model was further extended to allow the estimation of reciprocal, directed evolution between traits [@bartoszek2012phylogenetic], a method that has been implemented in the mvslouch package [@bartoszek2024fast]. This approach aligns closest with our goals, however the current maximum-likelihood implementation only allows for Gaussian traits, suffers from estimation problems, and lacks flexibility (such as accounting for latent factors, measurement error, missing data, repeated measures, etc.), which is why what motivated us to pursue a Bayesian implementation in the Stan probabilistic programming language [@Carpenter2017].

@@

The limitations of current PCMs can be overcome by synthesizing their respective
strengths into a single generalizable framework for dynamic phylogenetic causal
inference, applicable to both discrete and continuous traits. We have attempted
to do so through the development of what we refer to simply as generalized
dynamic phylogenetic models (GDPMs) of trait coevolution. Our GDPM is driven by a 
multivariate stochastic differential equation, similar to a multivariate 
Ornstein-Uhlenbeck (OU) process (see section 2.4. for an overview
of the basic OU model) with greater model flexibility and complexity than is 
permitted by current implementations. Our approach arose from two complementary
goals: first, to extend previous Bayesian implementations of the multivariate OU
model such that the optimal trait value $\theta$ is updated dynamically (rather
than assuming piecewise constant, see 2.4.). Second, we 
wanted a general form for the optimal trait value $\theta$ that could be used to
assess directionality (X → Y vs. Y → X) and contingencies (X, then Y) in 
evolution, akin to Pagel's [-@Pagel1994] method and existing OU methods [@bartoszek2012phylogenetic; @bartoszek2024fast] but without restrictions on the
type or number of traits, and with more flexibility to account for the realities of comparative data.

@@

where $\textbf{A}_{\text{\#}}=(\textbf{A}\otimes\textbf{I})+(\textbf{I}\otimes\textbf{A})$, 
with $\otimes$ denoting the Kronecker product. $\textbf{I}$ is an identity 
matrix, $\text{row()}$ is an operation that takes elements of a matrix row-wise
and puts them in a column vector, and $\text{irow()}$ is the inverse of the row
operation [@Driver2017]. In the OU process, $\frac{dy}{dt}=0$ when $y=\theta$ 
(Box \autoref{box:ou-model}). Equivalently, we can calculate the equilibrium 
value $\theta$ for each trait $\eta_i$ in the GDPM as

@@

Note that the vector $\theta_{\eta}$ does not equal the time asymptotic trait 
values for the system as a whole ($\textbf{b}_{\Delta\infty}$), which 
are determined by

@@

\begin{algorithm}
  \caption{Implementation of the generalized dynamic phylogenetic model}\label{alg:alg}
  \DontPrintSemicolon
  \For{$n \in N$}{
    Divide the evolutionary history of $n$ into $S$ segments, where each 
    segment starts with a parent node and ends with a child node or tip.\;
    Calculate the length of each segment $s \in S$, i.e., duration of time 
    between the parent and child nodes.\;
    Initialize the ancestral trait values $\eta_0$.\;
    \For{$s \in S$}{
      Solve for $\eta(s)$:
      \begin{equation}
        \nonumber
        \eta(s) = e^{\textbf{A}(t(s))}\eta(s-1)\textbf{A}^{-1}[e^{\textbf{A}
        (t(s)-t(s-1))}]\textbf{b} + \int_{t(s-1)}^{t(s)}e^{\textbf{A}
        (t(s-1)-t(s))}\textbf{G}dW(t(s))
      \end{equation}
      \;
    }
  }
\end{algorithm}

```{r GDPM-algorithm, echo=FALSE, fig.align='center', fig.cap='\\textit{The GDPM algorithm.} Example of how our phylogenetic algorithm works (see \\autoref{alg:alg} for formal structure). The evolutionary history of each tip (A-E) is described as a time series starting at \\(t(s)\\) and comprised of a set of segments \\(S\\). The OU process runs over each segment in order, where the initial trait values for any segment \\(s\\) are the terminal values from \\(s-1\\). Each sequence is a combination of shared segments (where tips are evolving together) and unshared segments (where tips are evolving independently). For example, S(A) and S(B) start with the same trait values because they both descend directly from the parent segment S(A,B). But further evolution along S(A) and S(B) happens independently. Dashed lines indicate segment split points.', out.width='100%'}
knitr::include_graphics("figures/dpm_algorithm.png")
```

@@

To aid researchers in applying the GDPM to their own data, we have developed the 
`coevolve` R package (https://github.com/ScottClaessens/coevolve). This 
fully documented package provides a user-friendly interface for applying GDPMs, 
with simple functions for generating GDPM Stan code and data lists, fitting GDPMs
to data using the `cmdstanr` package [@R_cmdstanr], and post-processing and 
plotting model results.

@@

We used the `SBC` R package [@Modrak2023] to run our simulation-based 
calibration exercise. We generated 500 simulated datasets from the prior ($N$ =
20, one continuous trait and one binary trait) and fitted the GDPM to these
simulated datasets. We then determined whether the posterior ranks from the 
fitted models were uniformly distributed by plotting, for several key model 
parameters and the log-likelihood, the differences between the empirical 
cumulative distribution function and a perfectly uniform cumulative distribution
function. If posterior ranks are perfectly uniformly distributed as expected, 
this should result in a flat line, though some variation from flat is to be 
anticipated. Ideally, we would expect to see the lines falling within the 
expected deviations at the 95% level (blue ovals). Indeed, this is what we find
for our model (\autoref{fig:sbc}), suggesting that the Stan program is 
well-calibrated.

@@

Brain size, both in absolute terms and relative to body size, varies 
tremendously among taxa, with primates as a group having larger brains than 
other mammals, and humans having the largest brains of all primates [@Boddy2012;
@Isler2008; @Miller2019; @Smaers2021]. This has motivated a large number of 
comparative phylogenetic analyses on correlates of brain size, typically framed
as testing social or ecological benefits or constraints for evolving larger 
brains [@DeCasien2017; @Dunbar2017; @Isler2009]. However, many have recognized 
that this approach has reached an impasse, as different hypotheses can be 
supported depending on the specific dataset [@Powell2017; @Wartel2019] due to 
variable patterns of missingness and high collinearity among predictors (e.g. 
diet, social group size, and life history), and reciprocal causation is inherent
in most theoretical models (e.g. larger brains allow access to better diet which
removes energetic constraints on larger brains) [@Isler2014]. Add the fact that
Pagel's discrete method has to our knowledge never been applied to brain size 
due to its continuous nature, and this field is virtually predisposed to the use
of GDPMs. The issues and innovations highlighted here should also generalize to 
many other morphological and socio-ecological traits.

To illustrate the potential of GDPMs to analyze brain size evolution, we used the
dataset of @DeCasien2017, augmented by life-history variables from 
@Herculano2019, and a consensus phylogeny from 10ktrees [@Arnold2010]. We 
contrasted a typical phylogenetic multiple regression (Fig. 5a), in which all 
predictors are interpreted as independent causes of brain size, with a latent 
factor model (Fig. 5b), which was derived in a bottom-up way using exploratory 
graph analysis (see Supplementary Materials). Note that instead of modeling 
absolute brain size and adjusting for body size as is typical for phylogenetic
regression (Fig. 5a), the model in Fig. 5b directly modeled relative brain size
as the allometric slope for the brain-body relationship; we also accounted for
variable allometric relationships using random intercepts and slopes at the
species level, thus allowing species to differ in their brain-body allometry
[@Smaers2021]. Note also that this GDPM used Bayesian imputation rather than 
removing species with missing data and losing valuable information and 
statistical power. In contrast to the approach in Fig. 5a, which compares 
independent causes and might conclude that "primate brain size is predicted by
diet but not sociality" [@DeCasien2017], the causal structure in Fig 5b groups
diet, social group size, and relative brain size into a single latent factor, 
termed here "niche complexity", with another latent factor, "life-history pace",
with loadings from the variables body size, age at maturity, and longevity. 
These latent factor groupings are not surprising, given that much work in 
primate socio-ecology has emphasized connections between diet and group size
[@Sterck1997; @Wrangham1980], and both are consistently found to correlate with
relative brain size [@DeCasien2017; @Dunbar2017]; the integration of life 
history traits and body size is also well established [@Healy2019; @Wright2019].

A GDPM analyzing the coevolution of these two latent factors (see Supplementary
Materials) showed that evolutionary changes in life-history pace (LH) cause 
evolutionary changes in niche complexity (NC), but not vice versa (Fig. 5c). In
other words, evolution of larger size and slower life history was inferred to 
precede increases in relative brain size, group size, and frugivory. This is 
consistent with the observations that repeated changes in size characterized
mammalian evolution as lineages radiated into new niches [@Pagel2022] and 
changes in body size associated with new niches were arguably primarily 
responsible for changes in relative brain size [@Smaers2021]. Further, debates
about the relative importance of diet and sociality in brain size evolution 
[@DeCasien2017; @Dunbar2017] seem to be moot as both are sides of the same 
socio-ecological coin intimately tied to relative brain size [see also 
@Powell2017; @Wartel2019]. Many extensions of this model are possible, and more
theoretically motivated causal models should be explored to further advance our
understanding of brain size evolution.
